animbook: Visualizing changes in performance measures and demographic affiliations using animation

### 2023-10-28 animbook-journal.Rmd
second version of the paper

The structure of this paper is as follows. Section @NYT explains the animation in The New York Times article and why it is relevant to the problem of studying zombie companies. The next section describes the expected data format, and how it is processed. Following this, is an explanation of available the animation tools, and how they are employed for this problem.  A  section on the functions of the package and how the visualization is designed, demonstrate how the data is mapped to the animation. The last two sections illustrate the usage of the package and applications to company performance and changing political allegiance.

@@

```{r nyt, fig.cap="Screenshot of the New York Times animation, that is the motivation for this visualization package.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page"}
knitr::include_graphics("figures/NYT.jpg")
```

@@

The principle underlying the animation is called persistence of vision (@animation-mechanic). When an image disappears, the brain will retain the previous images for a brief period of time. It is this slight period of retention that allows humans to separate sequential images. If this is seen in quick succession, it will appear as if the objects are in motion. This is illustrated in Figure \@ref(fig:animation-diagram). There are multiple ways to create an animation in the R environment (@r), including packages like `gganimate` and `plot`.

@@

  In the context of data, as shown in Figure \@ref(fig:animated-diagram), observations are positioned at specific points in time. The further the distance between these points, the less smooth the animation becomes. This issue can be eliminated by interpolating additional points in between the observations. In the `gganimate` package, the interpolation is achieved using the `tweenr` package (@tweenr), while in `plotly`, it utilizes d3.interpolate (@d3js).

@@

  Original data:

```{r}
dbl_change |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  reframe(id, time, values, gp) |> 
  arrange(gp, id)
```

  1. Ranking by year.

```{r}
rank <- anim_prep(dbl_change, id, values, time, color = gp)

rank$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  rename(gp = color) |> 
  select(-frame) |> 
  arrange(gp, id)
```

  2. Fix bins relative to absolute values by year.

```{r}
absolute <- anim_prep(dbl_change, id, values, time, color = gp, scaling = "absolute")

absolute$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |>  
  rename(gp = color) |> 
  select(-frame) |> 
  arrange(gp, id)
```

  3. Ranking by year within a group.

```{r}
rank_group <- anim_prep(dbl_change, id, values, time, color = gp, group_scaling = gp)

rank_group$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  rename(gp = group) |> 
  select(-c(frame, color)) |> 
  arrange(gp, id)
```

  4. Fix bins relative to absolute values by year within a group.

@@

absolute_group$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  rename(gp = group) |> 
  select(-c(frame, color)) |> 
  arrange(gp, id)
```

  For the first and third scaling methods, group splitting is executed using the `quantiles()` and `cut()` functions. The `quantile` function from the `stats` R package (@stats) takes a numeric vector and outputs the corresponding quantiles to the given probabilities. The output from the `quantile` function is then used as the `breaks` argument for the cut function that is part of the base R packages.

  In contrast, the second and fourth scaling methods calculate the quantile based on the absolute values scales. The default approach is to break the group equally using the `seq()` function. The `seq()` function takes input values from the minimum and maximum values and increments by equal steps depending on the number of groups of interest. Additionally, the users have the option to specify the breaks themselves if they choose to do so.

@@

  For the `anim_prep()` function, the arguments are as follows:

- data: A data frame containing the data to be prepared for visualization.
- id: The column name that represents the unique identifier variable.
- values: The column name that contains the numeric values to be visualized.
- time: The column name represents the time variable.
- label: A vector of labels to be used for the y-axis in the visualization.
- ngroup:	The number of groups or categories to create for scaling values.
- breaks: A vector of breaks for creating bins.
- group_scaling: The column name that represents the grouping variable.
- color: The column name to be used in ggplot2::aes() for the plot function.
- time_dependent: Logical. Should the visualization be time-dependent? Default is TRUE.
- scaling: The scaling method to be used; "rank" or "absolute."
- runif_min: The minimum value for random addition to frame numbers.
- runif_max: The maximum value for random addition to frame numbers.

  For the `anim_prep_cat()` function, the arguments are as follows:

- data: A data frame containing the data to be prepared for visualization.
- id: The column name that represents the unique identifier variable.
- values: The column name that contains the numeric values to be visualized.
- time: The column name represents the time variable.
- label: A vector of labels to be used for the y-axis in the visualization.
- order: A vector of order for sorting the category values.
- color: The column name to be used in ggplot2::aes() for the plot function.
- time_dependent: Logical. Should the visualization be time-dependent? Default is TRUE.
- runif_min: The minimum value for random addition to frame numbers.
- runif_max: The maximum value for random addition to frame numbers.

@@

Step 1: Prepare the data

@@

Step 2: Plotting the data

```{r, echo=TRUE, fig.width=8, fig.align='center', out.width="100%", layout = "l-page"}
p <- wallaby_plot(object = animbook,
             group_palette = RColorBrewer::brewer.pal(9, "Set1"),
             shade_palette = c("#737373", "#969696", "#BDBDBD","#D9D9D9","#D9D9D9","#D9D9D9"),
             rendering = "ggplot",
             subset = "top",
             relation = "one_many",
             total_point = NULL)

@@

Step 3: Animating the plot

@@

gganimate::anim_save("figures/catchange.gif", p2)
```

@@

@@

p <- wallaby_plot(accounting,
                  group_palette = RColorBrewer::brewer.pal(9, "Set1"),
                  shade_palette = c("#737373", "#969696", "#BDBDBD","#D9D9D9","#D9D9D9","#D9D9D9"),
                  subset = "bottom",
                  relation = "many_one",
                  height = 1,
                  size = 2,
                  width = 100,
                  total_point = 1000)

@@

```{r osiris-figure, fig.cap = "The animate visualization shows the companies that exited the market.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/animation-exit.png")
```

```{r osiris-gif, out.extra="class = 'gif'", fig.cap = "The animate visualization shows the companies that exited the market.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/exit.gif")
```

@@

## Voter behaviour

@@

p_voter <- wallaby_plot(object = voter,
                  group_palette = RColorBrewer::brewer.pal(9, "Set1"),
                  shade_palette = c("#737373", "#969696", "#BDBDBD","#D9D9D9","#D9D9D9","#D9D9D9"),
                  rendering = "ggplot",
                  subset = "top",
                  relation = "one_many",
                  height = 1,
                  size = 2,
                  width = 100,
                  total_point = 1000)

@@

```{r voter-figure, fig.cap = "The animate visualization shows how does the top party perform in keeping the old  voters for different genders.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/animation-voter.png")
```

```{r voter-gif, out.extra="class = 'gif'", fig.cap = "The animate visualization shows how does the top party perform in keeping the old  voters for different genders.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/voter.gif")
```

@@

Beginning with inspiration from the New York Times articles, this package provides tools to facilitate the communication of complex data to general audiences. In the current version of `animbook`, the feature to have different speed for each observation have not yet been implemented. This feature will introduce a new dimension that allows the users to explore the data further.

@@

This paper is based on the 0.0.0.9 version of the `animbook` package. This version is available on [GitHub](https://github.com/KrisanatA/animbook).

>>>>

The structure of this paper is as follows. The section explains the animation in The New York Times article and why it is relevant to the problem of studying zombie companies. The next section describes the expected data format. Following this is an explanation of available animation tools, and how they are employed for this problem. A section on the functions of the package and how the visualization is designed, demonstrates how the data is mapped to the animation. The last two sections illustrate the usage of the package and applications to company performance and changing political allegiance.

@@

```{r nyt, fig.cap="Screenshot of the New York Times animation, that is the motivation for this visualization package.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page"}
knitr::include_graphics("figures/NYT.png")
```

@@

The principle underlying the animation is called persistence of vision (@animation-mechanic). When an image disappears, the brain will retain the previous images for a brief period of time. It is this slight period of retention that allows humans to separate sequential images. If this is seen in quick succession, it will appear as if the objects are in motion. This is illustrated in Figure \@ref(fig:animation-diagram). There are multiple ways to create an animation in the R environment (@r), including packages like `gganimate` and `Plotly`.

@@

  In the context of data, as shown in Figure \@ref(fig:animated-diagram), observations are positioned at specific points in time. The further the distance between these points, the less smooth the animation becomes. This issue can be eliminated by interpolating additional points in between the observations. In the `gganimate` package (@gganimate), the interpolation is achieved using the `tweenr` package (@tweenr), while in `plotly`, it utilizes d3.interpolate (@d3js).

@@

```{r}
original <- dbl_change |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  reframe(id, time, values, gp) |> 
  arrange(gp, id)

rank <- anim_prep(dbl_change, id, values, time, color = gp)

rank_data <- rank$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  rename(gp = color,
         rank = qtile) |> 
  select(-frame) |> 
  arrange(gp, id)

absolute <- anim_prep(dbl_change, id, values, time, color = gp, scaling = "absolute")

absolute_data <- absolute$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |>  
  rename(gp = color,
         absolute = qtile) |> 
  select(-frame) |> 
  arrange(gp, id)

rank_group <- anim_prep(dbl_change, id, values, time, color = gp, group_scaling = gp)

rank_group_data <- rank_group$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  rename(gp = group,
         rank_group = qtile) |> 
  select(-c(frame, color)) |> 
  arrange(gp, id)
```

@@

absolute_group_data <- absolute_group$data |> 
  filter(id %in% c(1, 14, 21, 100, 106, 148)) |> 
  rename(gp = group,
         absolute_group = qtile) |> 
  select(-c(frame, color)) |> 
  arrange(gp, id)
```

```{r}
original |> 
  left_join(rank_data, by = c("id", "time", "gp")) |> 
  left_join(rank_group_data, by = c("id", "time", "gp")) |> 
  left_join(absolute_data, by = c("id", "time", "gp")) |> 
  left_join(absolute_group_data, by = c("id", "time", "gp")) |> 
  reframe(id, time, gp, values, rank, rank_group, absolute, absolute_group)
```

  1. Ranking by year. (rank)
  2. Ranking by year within a group. (rank_group)
  3. Fix bins relative to absolute values by year. (absolute)
  4. Fix bins relative to absolute values by year within a group. (absolute_group)

  For the first and second scaling methods, group splitting is executed using the `quantiles()` and `cut()` functions. The `quantile` function from the `stats` R package (@stats) takes a numeric vector and outputs the corresponding quantiles to the given probabilities. The output from the `quantile` function is then used as the `breaks` argument for the cut function that is part of the base R packages.

  In contrast, the third and fourth scaling methods calculate the quantile based on the absolute values scales. The default approach is to break the group equally using the `seq()` function. The `seq()` function takes input values from the minimum and maximum values and increments by equal steps depending on the number of groups of interest. Additionally, the users have the option to specify the breaks themselves if they choose to do so.

@@

  In the `anim_prep()` and `anim_prep_cat()`, there are required arguments that need to be specified. The lists are as follows:

  - data: A data frame containing the data to be prepared for visualization.
  - id: The column name that represents the unique identifier variable.
  - values: The column name that contains the numeric values to be visualized.
  - time: The column name represents the time variable.

  As shown previously, there are a total of four different scaling for the `anim_prep()` function. Both the use can customization of these scaling can be set using the following arguments:

  - ngroup:	The number of groups or categories to create for scaling values.
  - breaks: A vector of breaks for creating bins.
  - group_scaling: The column name that represents the grouping variable.
  - scaling: The scaling method to be used; "rank" or "absolute."

  For the `anim_prep_cat()` function, the arguments that allowed the user to order the category are by using this argument:

  - order: A vector of order for sorting the category values.

  Then, for further customization regarding how the final visualization looks. The following arguments can be adjusted:

  - label: A vector of labels to be used for the y-axis in the visualization.
  - color: The column name to be used in ggplot2::aes() for the plot function.
  - time_dependent: Logical. Should the visualization be time-dependent? Default is TRUE.
  - runif_min: The minimum value for random addition to frame numbers.
  - runif_max: The maximum value for random addition to frame numbers.

@@

In this section, the package usage will be demonstrated using the `cat_change` dataset included in this package. This dataset is simulated data that has changed from category A to E between two time points. Since this data is already in the categorized data structure, the `anim_prep_cat()` function is utilized.

  The output produced from this package is referred to as the `animbook` object. This contained a list of data and settings that hold prepared data for visualization, along with a list of settings used in the plot function.

@@

  To obtain a `ggplot` object for the animation function, the `wallaby_plot` is used. For further customization, the `shade_palette`, `subset`, and `relation` arguments are applied to define the final appearance of the animation.

```{r, echo=TRUE, fig.width=8, fig.align='center', out.width="100%", layout = "l-page"}
p <- wallaby_plot(object = animbook,
             group_palette = RColorBrewer::brewer.pal(9, "Set1"),
             shade_palette = c("#737373", "#969696", "#BDBDBD",
                               "#D9D9D9","#D9D9D9","#D9D9D9"),
             rendering = "ggplot",
             subset = "top",
             relation = "one_many",
             total_point = NULL)

@@

  Once the `ggplot` object is obtained, the `anim_animate()` function can be used to transform it into a `gganimate` object. This then can be passed onto the `gganimate::animate()` function for rendering.

@@

gganimate::anim_save("figures/catchange.gif", p2, height = 8, fps = 30,
                     nframes = 400, width = 9, units = "in", res = 150)
```

@@

  This is an example of how the users can apply the three-step process in creating an animation plot using the `animbook` package.

@@

p <- wallaby_plot(accounting,
                  group_palette = RColorBrewer::brewer.pal(9, "Set1"),
                  shade_palette = c("#737373", "#969696", "#BDBDBD",
                                    "#D9D9D9","#D9D9D9","#D9D9D9"),
                  subset = "bottom",
                  relation = "many_one",
                  height = 1,
                  size = 2,
                  width = 100,
                  total_point = 1000)

@@

```{r osiris-figure, fig.cap = "The animate visualization shows the companies that exited the market. There are more United States companies that fall down into a not listed group compared to Japanese companies.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/animation-exit.png")
```

```{r osiris-gif, out.extra="class = 'gif'", fig.cap = "The animate visualization shows the companies that exited the market. There are more United States companies that fall down into a not listed group compared to Japanese companies.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/exit.gif")
```

@@

## Voter behavior

@@

p_voter <- wallaby_plot(object = voter,
                  group_palette = RColorBrewer::brewer.pal(9, "Set1"),
                  shade_palette = c("#737373", "#969696", "#BDBDBD",
                                    "#D9D9D9","#D9D9D9","#D9D9D9"),
                  rendering = "ggplot",
                  subset = "top",
                  relation = "one_many",
                  height = 1,
                  size = 2,
                  width = 100,
                  total_point = 1000)

@@

```{r voter-figure, fig.cap = "The animate visualization shows how does the top party perform in keeping the old voters for different genders. Most voters remain loyal to the party, but a small fraction of voters with roughly equal male-to-female ratio switch primarily to the other major party.", fig.width=8, fig.align='center', out.width="100%", layout = "l-page", eval=knitr::is_latex_output()}
knitr::include_graphics("figures/animation-voter.png")
```

```{r voter-gif, out.extra="class = 'gif'", fig.cap = "The animate visualization shows how does the top party perform in keeping the old  voters for different genders. Most voters remain loyal to the party, but a small fraction of voters with roughly equal male-to-female ratio switch primarily to the other major party.", eval=knitr::is_html_output()}
knitr::include_graphics("figures/voter.gif")
```

@@

Beginning with inspiration from the New York Times articles, this package provides tools to facilitate the communication of complex data to general audiences. In the current version of `animbook`, the feature to have different speed for each observation have not yet been implemented. This feature will introduce a new dimension that allows the users to explore the data further. Additionally, both new animated and static plots could also be added, as seen in the New York Times article (@the_new_york_time), to compare the demographics at a specific point in time. On top of that, the `cli` package (@cli) could be used for a more aesthetically pleasing command-line interface.

@@

This paper is created using the `rjtools` packages (@rjtools) and is based on the 0.0.0.9 version of the `animbook` package. This version is available on [GitHub](https://github.com/KrisanatA/animbook).

I also acknowledge the use of ChatGPT (https://chat.openai.com/) to paraphrase the sentence.
